---
title: "Social Contact Matrices for STRIDE — USA (Wisconsin, Dane County) 2026"
author: "Willem L"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 6, fig.height = 4)
knitr::opts_knit$set(root.dir = '/Users/lwillem/opt/stride-751')
```

## Preamble

This report prepares USA-specific social contact matrices for STRIDE. Age-specific contact rates by location (household, school, workplace, community) are derived from the Prem et al. (2020) synthetic contact matrices for the US, and conditioned on the school enrolment, employment and household structure observed in a synthetic Wisconsin (Dane County) population. The resulting age-specific contact rates are exported as an XML file for use by the STRIDE simulator.

## Configuration parameters

```{r params}
# load 'contactdata' package
suppressPackageStartupMessages(library('contactdata'))

# load simid.rtools package
# require(devtools)
# devtools::install_github("lwillem/simid_rtools", force = F, quiet = T)
library('simid.rtools')

# select country with ISO2 code
sel_country <- 'US'

# select population file
pop_file <- 'data/pop_usa_wisconsin_dane474k_c1000.csv'

# set output tag
cdata_tag    <- paste0('contact_matrix_usa_conditional')

# set output directory
output_dir <- smd_file_path('sim_output')
file_name  <- file.path(output_dir, cdata_tag)
```

## Helper function: contact rates by location

```{r cnt-data-function}
# define help function to load social data from Prem et al. by location
get_cnt_data <- function(location, country) {
   cnt_matrix <- contact_matrix(country = country,
                                   location = location,
                                   data_source = c("2020"))
    cnt_count <- rowSums(cnt_matrix)
    all_ages <- 0:94
    approx(x = seq(0, 75, 5),
           y = cnt_count,
           xout = all_ages,
           method = 'constant',
           rule = 2:2)$y
}
```

## Load and process data

### Contact data by location

```{r load-contact-data}
# get data by location ----
cnt_all       <- get_cnt_data("all", country = sel_country)
cnt_home      <- get_cnt_data("home", country = sel_country)
cnt_workplace <- get_cnt_data("work", country = sel_country)
cnt_school    <- get_cnt_data("school", country = sel_country)
cnt_other     <- get_cnt_data("other", country = sel_country)
```

```{r check-contact-data}
# check
cnt_all_list <- contact_df_countries(countries = sel_country,
                                     location = "all",
                                     geographic_setting = c("all"),
                                     data_source = c("2020")
                )
sum(cnt_all_list$contact[cnt_all_list$age_from == "75_80"])
cnt_all[76] # note: index 1 is age age 0
```

### Additional contacts

We accumulate "additional contacts" for cases where the contact data assumes contacts at a location (e.g. school) that the population data does not support (e.g. no enrolment), or where the number of contacts at home outnumbers the number of household members.

```{r additional-contacts-init}
# define vector to accumulate "additional contact"
cnt_additional <- cnt_all * 0
```

### Demography data

```{r load-demography}
# load file
pop_usa <- read.csv(pop_file, header = T, sep = ',')

# rename work_id to workplace_id
names(pop_usa) <- gsub('work', 'workplace', names(pop_usa))

# set 'missing' to NA
pop_usa$school_id[pop_usa$school_id == 0] <- NA
pop_usa$workplace_id[pop_usa$workplace_id == 0] <- NA

# define age breaks for the age distribution
breaks_ages <- (0:95) - 0.5

# age distribution
age_counts <- hist(pop_usa$age, breaks = breaks_ages, plot = FALSE)$counts
```

### School enrolment

```{r school-enrolment}
age_counts_school <- hist(pop_usa$age[!is.na(pop_usa$school_id)], breaks = breaks_ages, plot = FALSE)$counts
age_distr_school <- age_counts_school / age_counts
age_distr_school[is.na(age_distr_school)] <- 0

# calculate number of contacts conditional on being at school
cnt_school_conditional      <- cnt_school * (7 / 5) # school is open 5 out of 7
cnt_school_conditional[3:5] <- cnt_school_conditional[6] # extrapolate behaviour of age 5 to age 2 to 4

# set contact rates for ages not enrolled in school to zero
cnt_school_conditional[age_distr_school == 0] <- 0

# define school contacts for ages not at school as "additional"
cnt_additional     <- cnt_school * (age_distr_school == 0)
cnt_additional[20] <- cnt_additional[21] # adjust artefact for age 19 (not enrolled in US data, but high number of contacts observed)
```

### Employment

```{r employment}
age_counts_workplace <- hist(pop_usa$age[!is.na(pop_usa$workplace_id)], breaks = breaks_ages, plot = FALSE)$counts
age_distr_workplace <- age_counts_workplace / age_counts
age_distr_workplace[is.na(age_distr_workplace)] <- 0

# define conditional number of contacts for all ages as the average of a selection of the (most) active population
workplace_ages <- 18:69
workplace_ages_select <- 30:49
cnt_workplace_conditional <- cnt_workplace * 0 # start with zero's
cnt_workplace_conditional[workplace_ages + 1]  <- mean(cnt_workplace[workplace_ages_select + 1]) # index = age + 1

# calculate number of contacts conditional on being at work
cnt_workplace_conditional <- cnt_workplace_conditional * (7 / 5)  # account for working 5 days out of 7
cnt_workplace_conditional <- cnt_workplace_conditional / mean(age_distr_workplace[workplace_ages_select + 1]) # account for employment rate
```

```{r workplace-size-check}
# optional: increase rates to account for small workplaces (n < mean number of contacts)?
workplace_size_count <- table(table(pop_usa$workplace_id))
# number of people in workplaces with <= 7 people
num_people_workplace_leq7 <- sum(workplace_size_count[1:7] * 1:7)
# proportional to number of workers
num_people_workplace_leq7 / sum(!is.na(pop_usa$workplace_id))
```

### Household contacts

```{r household-contacts}
# define household sizes
hh_size <- data.frame(table(pop_usa$household_id))
names(hh_size) <- c('household_id', 'household_size')

# add household info to population matrix
pop_usa_edit <- merge(pop_usa, hh_size)

# count number of households by member age and size
num_hh_age_size <- table(pop_usa_edit$age, pop_usa_edit$household_size)
num_age <- rowSums(num_hh_age_size)

# define matrix to represent number of hh contacts by hh size
mat_cnt_size <- matrix(rep(1:ncol(num_hh_age_size) - 1, length(num_age)),
                       ncol = ncol(num_hh_age_size),
                       byrow = T)

# assume fully connected households, get total number of contacts by age
hh_cnt_size <- num_hh_age_size * mat_cnt_size

# get mean number of contacts at home by age if household is fully connected
cnt_home_fully_connected <- rowSums(hh_cnt_size) / num_age

# impute/approximate missing ages
cnt_home_fully_connected <- approx(names(cnt_home_fully_connected),
                                   cnt_home_fully_connected,
                                   0:(length(cnt_home) - 1))$y

# set contacts with non-household-members as "additional"
cnt_additional <- cnt_additional + (cnt_home - cnt_home_fully_connected)
```

### Community

```{r community-contacts}
# start from cnt_other and add "additional"
cnt_other_adj <- cnt_other + cnt_additional
```

## Explore: conditional vs. unconditional contact rates

```{r explore-function}
# define function to explore (un)conditional contact rates
plot_conditional_contacts <- function(cnt_orig, cnt_conditional, pop_fraction, plot_main, xlim = c(0, 95)) {

  # define y_limit
  ylim <- c(0, max(c(cnt_orig, cnt_conditional)) * (4 / 3))

  # set margin
  par(mar = c(5, 5, 2, 5))

  # plot (un)conditional contact rates
  plot(cnt_conditional, xlim = xlim,
       ylim = ylim,
       main = plot_main,
       xlab = "age",
       ylab = "mean number of contacts")
  points(cnt_orig, col = 2)
  legend('topleft', c('unconditional and 7d average', 'conditional'), fill = 2:1)

  # set population fraction scaling factor
  if (!any(is.na(pop_fraction))) {
    frac_schale <- max(cnt_conditional) * 3 / 4
    lines(pop_fraction * frac_schale, col = 4)
    abline(h = frac_schale, lty = 3, col = 4)
    axis(4, seq(0, frac_schale, length.out = 11), labels = seq(0, 10, 1) / 10, las = 2, col = 4)
    mtext(paste('fraction enrolled in', plot_main), side = 4, padj = 4, col = 4, adj = 0.1)
  }
}
```

```{r explore-school, fig.cap='Conditional vs. unconditional contact rates, school'}
plot_conditional_contacts(cnt_school, cnt_school_conditional, age_distr_school, 'school', xlim = c(0, 22))
```

```{r explore-workplace, fig.cap='Conditional vs. unconditional contact rates, workplace'}
plot_conditional_contacts(cnt_workplace, cnt_workplace_conditional, age_distr_workplace, 'workplace')
```

```{r explore-household, fig.cap='Conditional vs. unconditional contact rates, household'}
plot_conditional_contacts(cnt_home, cnt_home_fully_connected, NA, 'household')
```

```{r explore-community, fig.cap='Conditional vs. unconditional contact rates, other/community'}
plot_conditional_contacts(cnt_other, cnt_other_adj, NA, 'other')
```

```{r explore-total, fig.cap='Contact rates, total'}
plot_conditional_contacts(cnt_all, cnt_all, NA, 'total')
```

## Save results

### Store as RData (for R)

```{r store-rdata}
# start with info on data and methods
cnt_data_meta <- list(data_source = 'USA social contact based on Prem et al (2017)',
                        method      = "Reported average number of contacts by age group, conditional on presence",
                        author      = Sys.info()['user'],
                        date        = format(Sys.time())
                        )

# add social contact data
social_cnt_data                     <- cnt_data_meta
social_cnt_data$regular_weekday     <- cnt_all
social_cnt_data$regular_weekend     <- cnt_all
social_cnt_data$household           <- cnt_home_fully_connected
social_cnt_data$school              <- cnt_school_conditional
social_cnt_data$workplace           <- cnt_workplace_conditional
social_cnt_data$community_weekday   <- cnt_other_adj
social_cnt_data$community_weekend   <- cnt_other_adj

save(social_cnt_data, file = paste0(file_name, '.RData'))
```

### Store as XML (for STRIDE)

```{r store-xml}
library('XML')

# create xml prefix
xml_prefix <- paste0(' This file is part of the Stride software [', format(Sys.time()), ']')

cnt_matrices_lib <- list(regular_weekday     = cnt_all,
                         regular_weekend     = cnt_all,
                         household           = cnt_home_fully_connected,
                         school              = cnt_school_conditional,
                         workplace           = cnt_workplace_conditional,
                         community_weekday   = cnt_other_adj,
                         community_weekend   = cnt_other_adj
                         )
# setup XML doc (to add prefix)
xml_doc <- newXMLDoc()

cnt_matrix_xml  <- newXMLNode("matrices", doc = xml_doc)
cnt_matrix_meta <- newXMLNode("metadata", parent = cnt_matrix_xml)
smd_listToXML(cnt_matrix_meta, cnt_data_meta)

for (i_context in 1:length(cnt_matrices_lib)) {

  # extract data
  cnt_context_name   <- names(cnt_matrices_lib)[i_context]
  cnt_context_values <- cnt_matrices_lib[[i_context]]

  # add data to XML
  cnt_context <- newXMLNode(cnt_context_name, parent = cnt_matrix_xml)
  for (i in 1:length(cnt_context_values)) {

    participant <- newXMLNode("participant", parent = cnt_context)

    part_age  <- newXMLNode("age", parent = participant)
    xmlValue(part_age) <- paste(i)

    contacts  <- newXMLNode("contacts", parent = participant)

    contact        <- newXMLNode("contact", parent = contacts)
    age            <- newXMLNode("age", parent = contact)
    xmlValue(age)  <- 'all'
    rate           <- newXMLNode("rate", parent = contact)
    xmlValue(rate) <- paste(cnt_context_values[i])
  }
}

# create filename for xml output
out_filename <- paste0(file_name, '.xml')

# save as XML,
# note: if we use an XMLdoc to include prefix, the line break dissapears...
# fix: http://r.789695.n4.nabble.com/saveXML-prefix-argument-td4678407.html
cat(saveXML(xml_doc, indent = TRUE, prefix = newXMLCommentNode(xml_prefix)), file = out_filename)
```

The final contact matrices for STRIDE are written to `r out_filename`.
